Precise temporal localisation of M/EEG effects with Bayesian generalised additive multilevel models

### 2025-08-22 manuscript.tex
pushing revised manuscript

\documentclass[
  doc,
  floatsintext,
  longtable,
  a4paper,
  nolmodern,
  notxfonts,
  notimes,
  donotrepeattitle,
  colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{apa7}

@@

\usepackage{longtable, booktabs, multirow, multicol, colortbl, hhline, caption, array, float, xpatch}
\setcounter{topnumber}{2}
\setcounter{bottomnumber}{2}
\setcounter{totalnumber}{4}
\renewcommand{\topfraction}{0.85}
\renewcommand{\bottomfraction}{0.85}
\renewcommand{\textfraction}{0.15}
\renewcommand{\floatpagefraction}{0.7}

@@

\abstract{Time-resolved electrophysiological measurements such as those
obtained through magneto- or electro-encephalography (M/EEG) offer a
unique window onto the neural activity underlying cognitive processes.
Researchers are often interested in determining whether and when these
signals differ across experimental conditions or participant groups. The
conventional approach involves mass-univariate statistical testing
across time and space, followed by corrections for multiple comparisons
such as cluster-based inference. While effective for controlling error
rates at the cluster-level, cluster-based inference comes with a
significant limitation: by shifting the focus of inference from
individual time points to clusters, it prevents drawing conclusions
about the precise onset or offset of observed effects. Here, we
introduce a \emph{model-based} approach for analysing M/EEG timeseries
such as event-related potentials (ERPs) or decoding performance over
time. Our approach leverages Bayesian generalised additive multilevel
models, providing posterior probabilities that an effect is above zero
(or above chance) at each time point, while naturally accounting for
temporal dependencies and between-subject variability. Using both
simulated and actual M/EEG datasets, we demonstrate that this approach
substantially outperforms conventional methods in estimating the onset
and offset of neural effects, yielding more precise and reliable
results. We provide an R package implementing the method and describe
how it can be integrated into M/EEG analysis pipelines using
MNE-Python.}

\keywords{EEG, MEG, cluster-based inference, simulation, multiple
comparisons, generalised additive models, mixed-effects
models, multilevel models, Bayesian statistics, brms}

@@

\par{   The authors have no conflicts of interest to disclose.    }
\par{Correspondence concerning this article should be addressed
to Ladislas Nalborczyk, Aix Marseille Univ, CNRS, LPL, 5 avenue
Pasteur, 13100
Aix-en-Provence, France, email: ladislas.nalborczyk@cnrs.fr}
}

@@

\setcounter{secnumdepth}{-\maxdimen} % remove section numbering

@@

The conventional approach involves mass-univariate statistical testing
through time and/or space followed by some form of correction for
multiple comparisons with the goal of maintaining the familywise error
rate (FWER) or the false discovery rate (FDR) at the nominal level
(e.g., 5\%). Cluster-based inference is the most common way of achieving
this sort of error control in the M/EEG literature, being the
recommended approach in several software programs (e.g.,
\texttt{EEGlab}, \citeproc{ref-delorme2004}{Delorme \& Makeig, 2004};
\texttt{MNE-Python}, \citeproc{ref-gramfort2013}{Gramfort, 2013}). While
effective for controlling error rates, cluster-based inference comes
with a significant limitation: by shifting the focus of inference from
individual datapoints (e.g., timesteps, sensors, voxels) to clusters, it
prevents the ability to draw precise conclusions about the
spatiotemporal localisation of such effects
(\citeproc{ref-maris2007}{Maris \& Oostenveld, 2007};
\citeproc{ref-sassenhagen2019}{Sassenhagen \& Draschkow, 2019}). As
pointed out by Maris \& Oostenveld (\citeproc{ref-maris2007}{2007}):
``there is a conflict between this interest in localized effects and our
choice for a global null hypothesis: by controlling the FA {[}false
alarm{]} rate under this global null hypothesis, one cannot quantify the
uncertainty in the spatiotemporal localization of the effect''. Even
worse, Rosenblatt et al. (\citeproc{ref-rosenblatt2018}{2018}) note that
cluster-based inference suffers from low spatial resolution: ``Since
discovering a cluster means that `there exists at least one voxel with
an evoked response in the cluster', and not that `all the voxels in the
cluster have an evoked response', it follows that the larger the
detected cluster, the less information we have on the location of the
activation.'' As a consequence, cluster-based inference is expected to
perform poorly for localising the onset of M/EEG effects; a property
that was later demonstrated in simulation studies (e.g.,
\citeproc{ref-rousselet_using_2025}{Rousselet, 2025};
\citeproc{ref-sassenhagen2019}{Sassenhagen \& Draschkow, 2019}).

To overcome the limitations of cluster-based inference, we introduce a
novel \emph{model-based} approach for precisely localising M/EEG effects
in time, space, and other dimensions. The proposed approach, based on
Bayesian generalised additive multilevel models, allows quantifying the
posterior probability of effects being above chance at the level of
timesteps, sensors, voxels, etc, while naturally taking into account
spatiotemporal dependencies present in M/EEG data. We compare the
performance of the proposed approach to well-established alternative
methods using both simulated and actual M/EEG data and show that it
significantly outperforms alternative methods in estimating the onset
and offset of M/EEG effects.

@@

The issues with multiple comparisons represent a common and
well-recognised danger in neuroimaging and M/EEG research, where the
collected data allows for a multitude of potential hypothesis tests and
is characterised by complex structures of spatiotemporal dependencies.
The probability of obtaining at least one false positive in an ensemble
(family) of \(m\) tests (i.e., the FWER) is computed as
\(1-\left(1-\alpha\right)^{m}\) (for \(m=10\) tests and \(\alpha=0.05\),
it is approximately equal to \(0.4\)). Different methods exist to
control the FWER (i.e., bring it back to \(\alpha\)). Most methods apply
a simple correction to series of \(p\)-values issued from univariate
statistical tests (e.g., t-tests). For instance, the Bonferroni
correction (\citeproc{ref-dunn1961}{Dunn, 1961}) consists in setting the
significance threshold to \(\alpha/m\), or equivalently, multiplying the
\(p\)-values by \(m\) and using the standard \(\alpha\) significance
threshold. This method is generally overconservative (i.e.,
under-powered) as it assumes statistical independence of the tests, an
assumption that is clearly violated in the context of M/EEG timeseries
characterised by massive spatiotemporal dependencies. Some alternative
methods aims at controlling the FDR, defined as the proportion of false
positive \emph{among positive tests} (e.g.,
\citeproc{ref-benjamini1995}{Benjamini \& Hochberg, 1995};
\citeproc{ref-benjamini2001}{Benjamini \& Yekutieli, 2001}). However, a
major limitation of both types of corrections is that they do not take
into account the spatial and temporal information contained in M/EEG
data.

A popular technique to account for spatiotemporal dependencies while
controlling the FWER is cluster-based inference
(\citeproc{ref-bullmore1999}{Bullmore et al., 1999};
\citeproc{ref-maris2007}{Maris \& Oostenveld, 2007}). A typical
cluster-based inference consists of two successive steps (for more
details on cluster-based inference, see for instance
\citeproc{ref-frossard2022}{Frossard \& Renaud, 2022};
\citeproc{ref-maris2011}{Maris, 2011}; \citeproc{ref-maris2007}{Maris \&
Oostenveld, 2007}; \citeproc{ref-sassenhagen2019}{Sassenhagen \&
Draschkow, 2019}). First, clusters are defined as sets of contiguous
timesteps, sensors, voxels, etc, whose activity, summarised by some test
statistic (e.g., a \(t\)-value), exceeds a predefined threshold (e.g.,
the 95th percentile of the parametric null distribution). Clusters are
then characterised by their height (i.e., maximal value), extent (number
of constituent elements), or some combination of both, for instance by
summing the statistics within a cluster, an approach referred to as
``cluster mass'' (\citeproc{ref-maris2007}{Maris \& Oostenveld, 2007};
\citeproc{ref-pernet2015}{Pernet et al., 2015}). Then, the null
hypothesis is tested by computing a \(p\)-value for each identified
cluster by comparing its mass with the null distribution of cluster
masses (obtained via permutation). As alluded previously, a significant
cluster is a cluster which contains \emph{at least one} significant
time-point. As such, it would be incorrect to conclude, for instance,
that the timestep of a significant cluster is the first moment at which
some conditions differ (\citeproc{ref-frossard2022}{Frossard \& Renaud,
2022}; \citeproc{ref-sassenhagen2019}{Sassenhagen \& Draschkow, 2019}).
In other words, because the inference is performed at the second step
(i.e., once clusters have been formed), no conclusion can be made about
individual datapoints (e.g., timesteps, sensors, etc).

@@

where \(h_{0}\) is typically \(0\) and parameters \(E\) and \(H\) are
set a priori (typically to 0.5 and 2, respectively) and control the
influence of the extend and height on the TFCE. Note that in practice,
this integral is approximated by a sum over small \(h\) increments.
Then, a \emph{p}-value for each timestep \(t\) is computed by comparing
its TFCE with the null distribution of TFCE values (obtained via
permutation). For each permuted signal, we keep the maximal value over
the whole signal for the null distribution of the TFCE. The TFCE
combined with permutation (assuming a large enough number of
permutations) has been shown to provide accurate FWER control (e.g.,
\citeproc{ref-pernet2015}{Pernet et al., 2015}). However, further
simulation work showed that cluster-based methods (including TFCE)
perform poorly in localising the onset of M/EEG effects (e.g.,
\citeproc{ref-rousselet_using_2025}{Rousselet, 2025};
\citeproc{ref-sassenhagen2019}{Sassenhagen \& Draschkow, 2019}).

@@

Scalp-recorded M/EEG signals capture neural activity originating from
various brain regions and are often contaminated by artifacts unrelated
to the cognitive processes under investigation. Consequently, analysing
M/EEG data necessitates methods that can disentangle task-relevant
neural signals from extraneous ``noise.'' A widely adopted technique for
this purpose is the estimation of event-related potentials (ERPs), which
are stereotyped electrophysiological responses time-locked to specific
sensory, cognitive, or motor events. Typically, ERPs are derived by
averaging EEG or MEG epochs across multiple trials aligned to the event
of interest (e.g., stimulus onset), thereby enhancing the
signal-to-noise ratio by attenuating non-time-locked activity. However,
this averaging approach has notable limitations: it assumes consistent
latency and amplitude across trials and is primarily suited for simple
categorical designs. Such assumptions may not hold in more complex
experimental paradigms, potentially leading to suboptimal ERP
estimations.

To overcome the limitations of simple averaging, several model-based
approaches for estimating ERPs have been proposed. These methods are
motivated by the observation that traditional ERP averaging is
mathematically equivalent to fitting an intercept-only linear regression
model in a simple categorical design without overlapping events
(\citeproc{ref-smith2014a}{N. J. Smith \& Kutas, 2014a}). In contrast to
simple averaging, regression-based approaches to ERP estimation offer
substantially greater flexibility. Notably, they allow for the modelling
of both linear and nonlinear effects of continuous predictors--such as
word frequency or age (e.g., \citeproc{ref-smith2014a}{N. J. Smith \&
Kutas, 2014a}, \citeproc{ref-smith2014b}{2014b};
\citeproc{ref-tremblay2014}{Tremblay \& Newman, 2014})--and enable the
disentangling of overlapping cognitive processes (e.g.,
\citeproc{ref-ehinger_unfold_2019}{Ehinger \& Dimigen, 2019};
\citeproc{ref-skukies_brain_2024}{Skukies et al., 2024};
\citeproc{ref-skukies_modelling_2021}{Skukies \& Ehinger, 2021}). One
widely used implementation of this approach is provided by the LIMO EEG
toolbox (\citeproc{ref-pernet2011}{Pernet et al., 2011}), which follows
a multi-stage analysis pipeline. First, a separate regression model is
fit for each datapoint (e.g., each time point and electrode) at the
individual participant level to estimate ERP responses. This is followed
by group-level statistical analyses of the resulting regression
coefficients, often accompanied by multiple comparison corrections or
cluster-based inference (for recent applied examples, see
\citeproc{ref-dunagan2025}{Dunagan et al., 2025};
\citeproc{ref-wuxfcllhorst2025}{Wüllhorst et al., 2025}).

Although this framework allows for the inclusion of a wide range of
predictors--both continuous and categorical, linear and nonlinear--it
still has important limitations. First, fitting separate models for each
datapoint ignores the spatiotemporal dependencies inherent in M/EEG
data, potentially reducing statistical power and interpretability.
Second, the subsequent group-level analyses typically do not account for
hierarchical dependencies, between-participant or between-stimulus
variability, which could otherwise be addressed through multilevel
modelling. Finally, because the output of this procedure is summarised
by cluster-based inference, its conclusions remain subject to the
limitations discussed in the previous section.

Beyond modelling nonlinear effects of continuous predictors on ERP
amplitudes, generalised additive models (GAMs) have been employed to
capture the temporal dynamics of ERPs themselves, effectively modelling
the shape of the waveform over time
(\citeproc{ref-abugaber2023}{Abugaber et al., 2023};
\citeproc{ref-baayen2018}{Baayen et al., 2018};
\citeproc{ref-meulman2015}{Meulman et al., 2015},
\citeproc{ref-meulman2023}{2023}). This approach allows for the
estimation of smooth, data-driven functions that characterise how neural
responses evolve over time, offering a flexible alternative to
traditional linear models. In the following section, we provide a brief
introduction to GAMs, highlighting their applicability to M/EEG time
series analysis and the advantages they offer over conventional methods.

@@

where \(\beta_{jk}\) is the weight (coefficient) associated with the
\(k\)th basis function \(b_{jk}()\) evaluated at the covariate value
\(x_{ij}\) for the \(j\)th smooth function \(f_{j}\). To clarify the
terminology at this point: \emph{splines} are functions composed of
simpler functions. These simpler functions are called \emph{basis
functions} (e.g., cubic polynomial, thin-plate) and the set of basis
functions is called a \emph{basis}. Each basis function gets its
coefficient and the resultant spline is the sum of these weighted basis
functions (Figure~\ref{fig-intro-gam}\ignorespaces A). Splines
coefficients are penalised (usually through the squared of the smooth
functions' second derivative) in a way that can be interpreted, in
Bayesian terms, as a prior on the ``wiggliness'' of the function
(\citeproc{ref-miller2025}{Miller, 2025}; \citeproc{ref-wood2017}{Wood,
2017a}). In other words, more complex (wiggly) basis functions are
automatically penalised.

@@

A detailed treatment of the technical underpinnings of GAMs is beyond
the scope of this article (see reference books such as
\citeproc{ref-hastie2017}{Hastie \& Tibshirani, 2017};
\citeproc{ref-wood2017}{Wood, 2017a}). However, it is worth emphasising
that GAMs have been successfully applied to a wide range of time series
data across the cognitive sciences, including pupillometry (e.g.,
\citeproc{ref-vanrij2019}{Rij et al., 2019}), articulography (e.g.,
\citeproc{ref-wieling2018}{Wieling, 2018}), speech formant dynamics
(e.g., \citeproc{ref-suxf3skuthy2021}{Sóskuthy, 2021}), neuroimaging
data (e.g., \citeproc{ref-dinga2021}{Dinga et al., 2021}), and
event-related potentials (e.g., \citeproc{ref-abugaber2023}{Abugaber et
al., 2023}; \citeproc{ref-baayen2018}{Baayen et al., 2018};
\citeproc{ref-meulman2015}{Meulman et al., 2015},
\citeproc{ref-meulman2023}{2023}). Their appeal for modelling M/EEG data
lies in their ability to flexibly capture the complex shape of ERP
waveforms without overfitting, through the use of smooth functions
constrained by penalisation. Recent extensions, such as distributional
GAMs (\citeproc{ref-rigby2005}{Rigby \& Stasinopoulos, 2005};
\citeproc{ref-umlauf2018}{Umlauf et al., 2018a}), allow researchers to
model not only the mean structure but also the variance (or scale) and
other distributional properties as functions of predictors--a feature
that has proven useful in modelling neuroimaging data (e.g.,
\citeproc{ref-dinga2021}{Dinga et al., 2021}). Moreover, hierarchical or
multilevel GAMs (\citeproc{ref-pedersen_hierarchical_2019}{E. J.
Pedersen et al., 2019}) provide a principled way to account for the
nested structure of M/EEG data (e.g., trials within participants),
enabling the inclusion of varying intercepts, slopes, and smoothers (as
illustrated in Figure~\ref{fig-intro-gam}\ignorespaces C-D). This
approach mitigates the risk of overfitting and reduces the influence of
outliers on smooth estimates (\citeproc{ref-baayen2020}{Baayen \& Linke,
2020}; \citeproc{ref-meulman2023}{Meulman et al., 2023}).

@@

Cluster-based permutation tests are widely used in M/EEG research to
identify statistically significant effects across time and space.
However, these methods have notable limitations, particularly in
accurately determining the precise onset and offset of neural effects.
To address these limitations, we developed a model-based approach
relying on Bayesian generalised additive multilevel models implemented
in \texttt{R} via the \texttt{brms} package
(\citeproc{ref-brms2017}{Bürkner, 2017}, \citeproc{ref-brms2018}{2018}).
We evaluated the performance of this approach against conventional
methods using both simulated and actual M/EEG data. Our findings
demonstrate that this method provides more precise and reliable
estimates of effect onsets and offsets, offering a valuable tool for
researchers seeking to enhance the temporal resolution of their
analyses.

\section{Methods}\label{methods}

\subsection{M/EEG data simulation}\label{meeg-data-simulation}

>>>>

\documentclass[
  doc,
  floatsintext,
  longtable,
  a4paper,
  nolmodern,
  notxfonts,
  notimes,
  colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{apa7}

@@

\usepackage{longtable, booktabs, multirow, multicol, colortbl, hhline, caption, array, float, xpatch}
\usepackage{subcaption}

\renewcommand\thesubfigure{\Alph{subfigure}}
\setcounter{topnumber}{2}
\setcounter{bottomnumber}{2}
\setcounter{totalnumber}{4}
\renewcommand{\topfraction}{0.85}
\renewcommand{\bottomfraction}{0.85}
\renewcommand{\textfraction}{0.15}
\renewcommand{\floatpagefraction}{0.7}

@@

\abstract{Time-resolved electrophysiological measurements such as those
obtained through magneto- and electroencephalography (M/EEG) offer a
unique window onto the neural activity underlying cognitive processes.
Researchers are often interested in determining whether and when these
signals differ across experimental conditions or participant groups. The
conventional approach involves mass-univariate statistical testing
across time and space followed by corrections for multiple comparisons
or some form of cluster-based inference. While effective for controlling
error rates at the cluster-level, cluster-based inference comes with a
significant limitation: by shifting the focus of inference from
individual time points to clusters, it prevents drawing conclusions
about the precise onset or offset of observed effects. Here, we present
a \emph{model-based} alternative for analysing M/EEG timeseries, such as
event-related potentials or time-resolved decoding accuracy. Our
approach leverages Bayesian generalised additive multilevel models,
providing posterior odds that an effect exceeds zero (or chance) at each
time point, while naturally accounting for temporal dependencies and
between-subject variability. Using both simulated and empirical M/EEG
datasets, we show that this approach substantially outperforms
conventional methods in estimating the onset and offset of neural
effects, yielding more precise and reliable estimates. We provide an
open-source R package implementing the method and describe how it can be
integrated into M/EEG analysis pipelines using MNE-Python. }

\keywords{EEG, MEG, cluster-based inference, multiple
comparisons, generalised additive models, mixed-effects
models, multilevel models, Bayesian statistics, brms}

@@

\par{   The authors have no conflicts of interest to disclose.    }
\par{Correspondence concerning this article should be addressed
to Ladislas Nalborczyk, Aix Marseille Univ, CNRS, LPL, 5 avenue
Pasteur, 13100
Aix-en-Provence, France, email: \href{mailto:ladislas.nalborczyk@cnrs.fr}{ladislas.nalborczyk@cnrs.fr}}
}

@@

\setcounter{secnumdepth}{3}

@@

The conventional approach involves mass-univariate statistical testing
through time and/or space followed by some form of correction for
multiple comparisons with the goal of maintaining the familywise error
rate (FWER) or the false discovery rate (FDR) at the nominal level
(e.g., 5\%). Cluster-based inference is the most common way of achieving
this sort of error control in the M/EEG literature, being the
recommended approach in several software programs (e.g.,
\texttt{EEGlab}, \citeproc{ref-delorme2004}{Delorme \& Makeig, 2004};
\texttt{MNE-Python}, \citeproc{ref-gramfort2013}{Gramfort, 2013}). While
effective for controlling error rates, cluster-based inference comes
with a significant limitation: by shifting the focus of inference from
individual datapoints (e.g., timesteps, sensors, voxels) to clusters, it
prevents the ability to draw precise conclusions about the
spatiotemporal localisation of such effects
(\citeproc{ref-maris2007}{Maris \& Oostenveld, 2007};
\citeproc{ref-sassenhagen2019}{Sassenhagen \& Draschkow, 2019}). As
pointed out by Maris \& Oostenveld (\citeproc{ref-maris2007}{2007}):
``there is a conflict between this interest in localized effects and our
choice for a global null hypothesis: by controlling the FA {[}false
alarm{]} rate under this global null hypothesis, one cannot quantify the
uncertainty in the spatiotemporal localization of the effect''. Even
worse, Rosenblatt et al. (\citeproc{ref-rosenblatt2018}{2018}) note that
cluster-based inference suffers from low spatial resolution: ``Since
discovering a cluster means that `there exists at least one voxel with
an evoked response in the cluster', and not that `all the voxels in the
cluster have an evoked response', it follows that the larger the
detected cluster, the less information we have on the location of the
activation.'' As a consequence, cluster-based inference is expected to
perform poorly for identifying the onset of M/EEG effects; a property
that was later demonstrated in simulation studies (e.g.,
\citeproc{ref-rousselet_using_2025}{Rousselet, 2025};
\citeproc{ref-sassenhagen2019}{Sassenhagen \& Draschkow, 2019}).

To overcome the limitations of cluster-based inference, we introduce a
novel \emph{model-based} approach for precisely localising M/EEG effects
in time, space, and other dimensions. The proposed approach, based on
Bayesian generalised additive multilevel models, allows quantifying the
posterior odds of effects being above chance at the level of timesteps,
sensors, voxels, etc, while naturally taking into account spatiotemporal
dependencies present in M/EEG data. We compare the performance of the
proposed approach to well-established alternative methods using both
simulated and actual M/EEG data and show that it significantly
outperforms alternative methods in estimating the onset and offset of
M/EEG effects.

@@

The issues with multiple comparisons represent a common and
well-recognised danger in neuroimaging and M/EEG research, where the
collected data allows for a multitude of potential hypothesis tests and
is characterised by complex structures of spatiotemporal dependencies.
The probability of obtaining at least one false positive in an ensemble
(family) of \(m\) tests (i.e., the FWER) is computed as
\(1-\left(1-\alpha\right)^{m}\) (for \(m=10\) independent tests and
\(\alpha=0.05\), it is approximately equal to \(0.4\)). Different
methods exist to control the FWER, that is, to bring it back to
\(\alpha\). Most methods apply a simple correction to series of
\(p\)-values issued from univariate statistical tests (e.g., t-tests).
For instance, the Bonferroni correction (\citeproc{ref-dunn1961}{Dunn,
1961}) consists in setting the significance threshold to \(\alpha/m\),
or equivalently, multiplying the \(p\)-values by \(m\) and using the
standard \(\alpha\) significance threshold. This method is generally
overconservative (i.e., under-powered) as it assumes statistical
independence of the tests, an assumption that is clearly violated in the
context of M/EEG timeseries characterised by massive spatiotemporal
dependencies. Some alternative methods aims at controlling the FDR,
defined as the proportion of false positive \emph{among positive tests}
(e.g., \citeproc{ref-benjamini1995}{Benjamini \& Hochberg, 1995};
\citeproc{ref-benjamini2001}{Benjamini \& Yekutieli, 2001}). However, a
major limitation of both types of corrections is that they do not take
into account the spatial and temporal information contained in M/EEG
data.

A popular technique to account for spatiotemporal dependencies while
controlling the FWER is cluster-based inference
(\citeproc{ref-bullmore1999}{Bullmore et al., 1999};
\citeproc{ref-maris2007}{Maris \& Oostenveld, 2007}). A typical
cluster-based inference consists of two successive steps (for more
details on cluster-based inference, see for instance
\citeproc{ref-frossard2022}{Frossard \& Renaud, 2022};
\citeproc{ref-maris2011}{Maris, 2011}; \citeproc{ref-maris2007}{Maris \&
Oostenveld, 2007}; \citeproc{ref-sassenhagen2019}{Sassenhagen \&
Draschkow, 2019}). First, clusters are defined as sets of contiguous
timesteps, sensors, voxels, etc, whose activity, summarised by some test
statistic (e.g., a \(t\)-value), exceeds a predefined threshold (e.g.,
the 95th percentile of the parametric null distribution). Clusters are
then characterised by their height (i.e., maximal value), extent (number
of constituent elements), or some combination of both, for instance by
summing the statistics within a cluster, an approach referred to as
``cluster mass'' (\citeproc{ref-maris2007}{Maris \& Oostenveld, 2007};
\citeproc{ref-pernet2015}{Pernet et al., 2015}). Then, the null
hypothesis is tested by computing a \(p\)-value for each identified
cluster by comparing its mass with the null distribution of cluster
masses (obtained via permutation). As alluded previously, a significant
cluster is a cluster which contains \emph{at least one} significant
time-point. As such, it would be incorrect to conclude, for instance,
that the timestep of a significant cluster is the first moment at which
some conditions differ (\citeproc{ref-frossard2022}{Frossard \& Renaud,
2022}; \citeproc{ref-sassenhagen2019}{Sassenhagen \& Draschkow, 2019}).
Because the inference is performed at the second step (i.e., once
clusters have been formed), no conclusion can be made about individual
datapoints (e.g., timesteps, sensors, etc).

@@

where \(h_{0}\) is typically \(0\) and parameters \(E\) and \(H\) are
set a priori (typically to 0.5 and 2, respectively) and control the
influence of the extend and height on the TFCE. In practice, this
integral is approximated by a sum over small \(h\) increments. Then, a
\emph{p}-value for each timestep \(t\) is computed by comparing its TFCE
with the null distribution of TFCE values (obtained via permutation).
For each permuted signal, we keep the maximal value over the whole
signal for the null distribution of the TFCE. The TFCE combined with
permutation (assuming a large enough number of permutations) has been
shown to provide accurate FWER control (e.g.,
\citeproc{ref-pernet2015}{Pernet et al., 2015}). However, further
simulation work showed that cluster-based methods (including TFCE)
perform poorly in localising the onset of M/EEG effects (e.g.,
\citeproc{ref-rousselet_using_2025}{Rousselet, 2025};
\citeproc{ref-sassenhagen2019}{Sassenhagen \& Draschkow, 2019}).

@@

Scalp-recorded M/EEG signals capture neural activity originating from
various brain regions and are often contaminated by artifacts unrelated
to the cognitive processes under investigation. Consequently, analysing
M/EEG data necessitates methods that can disentangle task-relevant
neural signals from extraneous ``noise.'' A widely adopted technique for
this purpose is the estimation of event-related potentials (ERPs), which
are stereotyped electrophysiological responses time-locked to specific
sensory, cognitive, or motor events. Typically, ERPs are derived by
averaging EEG or MEG epochs across multiple trials aligned to the event
of interest (e.g., stimulus onset), thereby enhancing the
signal-to-noise ratio by attenuating non-time-locked activity. However,
this averaging approach has notable limitations: it assumes consistent
latency and amplitude across trials and is primarily suited for simple
categorical designs. Such assumptions may not hold in more complex
experimental paradigms, potentially leading to suboptimal ERP
estimations (e.g., \citeproc{ref-smith2014a}{N. J. Smith \& Kutas,
2014a}).

To overcome the limitations of simple averaging, several model-based
approaches for estimating ERPs have been proposed. These methods are
motivated by the observation that traditional ERP averaging is
mathematically equivalent to fitting an intercept-only linear regression
model in a simple categorical design without overlapping events
(\citeproc{ref-smith2014a}{N. J. Smith \& Kutas, 2014a}). In contrast to
simple averaging, regression-based approaches to ERP estimation offer
substantially greater flexibility. Notably, they allow for the modelling
of both linear and nonlinear effects of continuous predictors, such as
word frequency or age (e.g., \citeproc{ref-smith2014a}{N. J. Smith \&
Kutas, 2014a}, \citeproc{ref-smith2014b}{2014b};
\citeproc{ref-tremblay2014}{Tremblay \& Newman, 2014}), and enable the
disentangling of overlapping cognitive processes (e.g.,
\citeproc{ref-ehinger_unfold_2019}{Ehinger \& Dimigen, 2019};
\citeproc{ref-skukies_brain_2024}{Skukies et al., 2024};
\citeproc{ref-skukies_modelling_2021}{Skukies \& Ehinger, 2021}). One
widely used implementation of this approach is provided by the LIMO EEG
toolbox (\citeproc{ref-pernet2011}{Pernet et al., 2011}), which follows
a multi-stage analysis pipeline. First, a separate regression model is
fit for each datapoint (e.g., each time point and electrode) at the
individual participant level to estimate ERP responses. This is followed
by group-level statistical analyses of the resulting regression
coefficients, often accompanied by corrections for multiple comparisons
or cluster-based inference (for recent applied examples, see
\citeproc{ref-dunagan2025}{Dunagan et al., 2025};
\citeproc{ref-wuxfcllhorst2025}{Wüllhorst et al., 2025}).

Although this framework allows for the inclusion of a wide range of
predictors--both continuous and categorical, linear and nonlinear--it
still has important limitations. First, fitting separate models for each
datapoint ignores the spatiotemporal dependencies inherent in M/EEG
data, potentially reducing statistical power and interpretability.
Second, the subsequent group-level analyses typically do not account for
hierarchical dependencies which could otherwise be addressed through
multilevel modelling. Finally, because the output of this procedure is
summarised by cluster-based inference, its conclusions remain subject to
the limitations discussed in the previous section.

Beyond modelling nonlinear effects of continuous predictors on ERP
amplitudes, GAMs have been employed to capture the temporal dynamics of
ERPs themselves, effectively modelling the shape of the waveform over
time (\citeproc{ref-abugaber2023}{Abugaber et al., 2023};
\citeproc{ref-baayen2018}{Baayen et al., 2018};
\citeproc{ref-meulman2015}{Meulman et al., 2015},
\citeproc{ref-meulman2023}{2023}). This approach allows for the
estimation of smooth, data-driven functions that characterise how neural
responses evolve over time, offering a flexible alternative to
traditional linear models. In the following section, we provide a brief
introduction to GAMs, highlighting their applicability to M/EEG time
series analysis and the advantages they offer over conventional methods.

@@

where \(\beta_{jk}\) is the weight (coefficient) associated with the
\(k\)th basis function \(b_{jk}()\) evaluated at the covariate value
\(x_{ij}\) for the \(j\)th smooth function \(f_{j}\). To clarify the
terminology at this point: \emph{splines} are functions composed of
simpler functions. These simpler functions are called \emph{basis
functions} (e.g., cubic polynomial, thin-plate) and the set of basis
functions is called a \emph{basis}. Each basis function is weighted by
its coefficient and the resultant spline is the sum of these weighted
basis functions (Figure~\ref{fig-intro-gam}\ignorespaces A). Splines
coefficients are penalised (usually through the square of the smooth
functions' second derivative) in a way that can be interpreted, in
Bayesian terms, as a prior on the ``wiggliness'' of the function
(\citeproc{ref-miller2025}{Miller, 2025}; \citeproc{ref-wood2017}{Wood,
2017a}). In other words, more complex (wiggly) basis functions are
automatically penalised.

@@

A detailed treatment of the technical underpinnings of GAMs is beyond
the scope of this article (see reference books such as
\citeproc{ref-hastie2017}{Hastie \& Tibshirani, 2017};
\citeproc{ref-wood2017}{Wood, 2017a}). However, it is worth emphasising
that GAMs have been successfully applied to a wide range of time series
data across the cognitive sciences, including pupillometry (e.g.,
\citeproc{ref-vanrij2019}{Rij et al., 2019}), articulography (e.g.,
\citeproc{ref-wieling2018}{Wieling, 2018}), speech formant dynamics
(e.g., \citeproc{ref-suxf3skuthy2021}{Sóskuthy, 2021}), neuroimaging
data (e.g., \citeproc{ref-dinga2021}{Dinga et al., 2021}), and
event-related potentials (e.g., \citeproc{ref-abugaber2023}{Abugaber et
al., 2023}; \citeproc{ref-baayen2018}{Baayen et al., 2018};
\citeproc{ref-meulman2015}{Meulman et al., 2015},
\citeproc{ref-meulman2023}{2023}). Their appeal for modelling M/EEG data
lies in their ability to flexibly capture the complex shape of ERP
waveforms without overfitting, through the use of smooth functions
constrained by penalisation. Recent extensions, such as distributional
GAMs (\citeproc{ref-rigby2005}{Rigby \& Stasinopoulos, 2005};
\citeproc{ref-umlauf2018}{Umlauf et al., 2018}), allow researchers to
model not only the mean structure but also the variance (or scale) and
other distributional properties as functions of predictors, a feature
that has proven useful in modelling neuroimaging data (e.g.,
\citeproc{ref-dinga2021}{Dinga et al., 2021}). Moreover, hierarchical or
multilevel GAMs (\citeproc{ref-pedersen_hierarchical_2019}{E. J.
Pedersen et al., 2019}) provide a principled way to account for the
nested structure of M/EEG data (e.g., trials within participants),
enabling the inclusion of varying intercepts, slopes, and smoothers (as
illustrated in Figure~\ref{fig-intro-gam}\ignorespaces C-D). This
approach mitigates the risk of overfitting and reduces the influence of
outliers on smooth estimates (\citeproc{ref-baayen2020}{Baayen \& Linke,
2020}; \citeproc{ref-meulman2023}{Meulman et al., 2023}).

@@

Cluster-based permutation tests are widely used in M/EEG research to
identify statistically significant effects across time and space.
However, these methods have notable limitations, particularly in
accurately determining the precise onset and offset of neural effects.
To address these limitations, we developed a model-based approach
relying on Bayesian generalised additive multilevel models implemented
in \texttt{R} via the \texttt{brms} package
(\citeproc{ref-brms2017}{Bürkner, 2017}, \citeproc{ref-brms2018}{2018}).
We evaluated the performance of this approach against conventional
methods using both simulated and actual M/EEG data. Our findings
demonstrate that this method provides more precise and reliable
estimates of effects' onset and offset than conventional approaches such
as cluster-based inference.

\section{Benchmarking with known ground
truth}\label{benchmarking-with-known-ground-truth}

\subsection{Methods}\label{methods}

\subsubsection{M/EEG data simulation}\label{meeg-data-simulation}

To assess the accuracy of group-level onset and offset estimation of our
proposed method, we simulated EEG with known onset and offset values.
Following the approach of Sassenhagen \& Draschkow
(\citeproc{ref-sassenhagen2019}{2019}) and Rousselet
(\citeproc{ref-rousselet_using_2025}{2025}), we simulated EEG data
stemming from two conditions, one with noise only, and the other with
noise + signal. As in previous studies, the noise was generated by
superimposing 50 sinusoids at different frequencies, following an
EEG-like spectrum (see code in the online supplementary materials and
details in \citeproc{ref-yeung2004}{Yeung et al., 2004}). As in
Rousselet (\citeproc{ref-rousselet_using_2025}{2025}), the signal was
generated from a truncated Gaussian distribution with an objective onset
at 160 ms, a peak at 250 ms, and an offset at 342 ms. We simulated this
signal for 250 timesteps between 0 and 0.5s, akin to a 500 Hz sampling
rate. We simulated data for a group of 20 participants (with variable
true onset) with 50 trials per participant and condition
(Figure~\ref{fig-eeg}). All figures and simulation results can be
reproduced using the \texttt{R} code available online at:
\url{https://github.com/lnalborczyk/brms_meeg}.
